---
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Reproducible Research: Peer Assessment 1
========================
Nasim Ahmed
=========================
2/10/2018
=========================

Introduction
==============
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

Data
=======
The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]:

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement data was taken in YYYY-MM-DD format

interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Assignment
=============
This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.

Throughout your report make sure you always include the code that you used to generate the output you present. When writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code. This assignment will be evaluated via peer assessment so it is essential that your peer evaluators be able to review the code for your analysis.

For the plotting aspects of this assignment, feel free to use any plotting system in R (i.e., base, lattice, ggplot2).

Fork/clone the GitHub repository created for this assignment. You will submit this assignment by pushing your completed files into your forked repository on GitHub. The assignment submission will consist of the URL to your GitHub repository and the SHA-1 commit ID for your repository state.

NOTE: The GitHub repository also contains the dataset for the assignment so you do not have to download the data separately.

Loading and preprocessing the data
====================================

```{r}
# I have downloaded the Zip file from the coursera provided link and unzip the file to load the dataset

activitydata <- read.csv("H:/Data Science Online Course/Reproducible Research/assingment/activity.csv", stringsAsFactors = FALSE)


activitydata$date <- as.POSIXct(activitydata$date, format = "%Y-%m-%d")

# The following  function (histogram) will calculate of the total number of steps taken in each day
activitydata <- data.frame(date = activitydata$date,
                           weekday = tolower(weekdays(activitydata$date)),
                           steps = activitydata$steps,
                           interval = activitydata$interval)

# Calculate the types of day
activitydata <- cbind(activitydata, daytype = ifelse(activitydata$weekday == "saturday" | activitydata$weekday == "sunday", "Weekend" , "Weekday"))
activity <- data.frame(date = activitydata$date, weekday = activitydata$weekday, daytype = activitydata$daytype,
                       interval = activitydata$interval, steps = activitydata$steps)

head(activity)

```

Total number of steps taken in each day
=======================================

```{r}
sumdata <- aggregate(activity$steps, by = list(activity$date), FUN = sum, na.rm = TRUE)
names(sumdata) <- c("date", "total")
head(sumdata)

# Average day data

library(dplyr)
Avgday <- activitydata %>% group_by(date) %>% summarize(total.steps = sum(steps, na.rm = TRUE), mean.steps = mean(steps, na.rm = TRUE)) 
summary(Avgday$total.steps)
summary(Avgday$mean.steps)

# Generate the histogram plot
hist(sumdata$total,
     breaks = seq(from = 0, to = 25000, by = 2500),
     col = "red", xlab = "Total Number of Steps", ylim = c(0, 30),
     main = "Total number of steps taken in each day\n(NA removed)")
```

Average daily activity pattern
==============================

```{r}

#Calculate 5 minutes interval and the average number of steps
mean_data <- aggregate(activity$steps, by = list(activity$interval), FUN = mean, na.rm = TRUE)
names(mean_data) <- c("interval", "mean")
head(mean_data)

# Time seris plot
plot(mean_data$interval, mean_data$mean, type = "l", col = "red", lwd = 2, xlab = "Interval[min]", ylab = "Avg number of steps",
     main = "Time series of the average number of steps per intervals\n(NA removed)")
maxposition <- which(mean_data$mean == max(mean_data$mean))
max_interval <- mean_data[maxposition, 1]

# Imputing missing values
mean(is.na(activitydata$steps))
sum(is.na(activitydata$steps))
```

Activity patterns between weekdays and weekends
===============================================
```{r}
# Search the NA position
NAposition <- which(is.na(activity$steps))
mean_vector <- rep(mean(activity$steps, na.rm = TRUE), times = length(NAposition))

# Now replace NAs by mean
activity[NAposition, "steps"] <- mean_vector
rm(mean_vector, NAposition)
head(activity)
newAvg <- activity %>% group_by(date) %>% summarize(total.steps = sum(steps, na.rm = T))

#Calculate each day total number of steps
sumdata <- aggregate(activity$steps, by = list(activity$date), FUN = sum)
names(sumdata) <- c("date", "total")
summary(Avgday$total.steps)
sd(Avgday$total.steps, na.rm = T)
summary (newAvg$total.steps)
sd(newAvg$total.steps, na.rm=T)

# Generate the histogram plot considering the total number of each day
hist(sumdata$total,
     breaks = seq(from = 0, to = 25000, by = 2500),
     col = "red", xlab = "Total Number of Steps", ylim = c(0, 30),
     main = "Total number of steps taken in each day\n(NA replaced by mean value)")


## Calculate weekday and weekend
library(lattice)
mean_data <- aggregate(activity$steps,by = list(activity$daytype, activity$weekday,  activity$interval), mean)
names(mean_data) <- c("daytype", "weekday", "interval", "mean")
#head(mean_data)

## Generate the plot

xyplot(mean ~ interval | daytype, mean_data, type = "l", lwd = 1, xlab = "interval", ylab = "Number of steps", layout = c(1, 2))
rm(mean_data)
```

